---
title: "QC & Filtering"
subtitle: "Assess and remove low-quality cells before downstream analysis"
categories: [scRNA-seq, Seurat, QC]
---

<span class="chapter-badge">scRNA-seq · Chapter 01</span>

## Overview

Raw single-cell RNA-seq data invariably contains low-quality cells — damaged cells
with compromised membranes, empty droplets, or multiplets (two or more cells captured
in one droplet). Retaining these artefacts distorts downstream clustering and
differential expression results.

**This chapter covers:**

1. Loading Cell Ranger output into a Seurat object
2. Computing per-cell QC metrics
3. Selecting filtering thresholds from QC distributions
4. Doublet detection with DoubletFinder
5. Saving the filtered object for normalization

---

## Setup

```{r}
#| label: setup

library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
```

---

## Load Data

```{r}
#| label: load-data
#| eval: false

# From Cell Ranger filtered output directory
counts <- Read10X(
  data.dir = "data/raw/pbmc3k/filtered_feature_bc_matrix/"
)

# --- OR from .h5 file ---
counts <- Read10X_h5("data/raw/pbmc3k/filtered_feature_bc_matrix.h5")

seu <- CreateSeuratObject(
  counts      = counts,
  project     = "PBMC3k",
  min.cells   = 3,    # retain genes detected in ≥ 3 cells
  min.features = 200  # retain cells with ≥ 200 genes detected
)

seu
```

::: {.callout-note}
`min.cells = 3` and `min.features = 200` are deliberately loose.
Stricter thresholds are applied after inspecting QC metric distributions.
:::

---

## Compute QC Metrics

```{r}
#| label: qc-metrics
#| eval: false

# Mitochondrial gene percentage — high % indicates damaged/dying cells
seu[["pct_mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")

# Ribosomal gene percentage — context-dependent, informative to track
seu[["pct_rb"]] <- PercentageFeatureSet(seu, pattern = "^RP[SL]")

# Preview metadata
head(seu@meta.data, 5)
```

---

## Visualise QC Distributions

```{r}
#| label: qc-violin
#| eval: false
#| fig-cap: "Per-cell QC metrics before filtering. Dashed lines indicate candidate thresholds."

VlnPlot(
  seu,
  features = c("nFeature_RNA", "nCount_RNA", "pct_mt"),
  ncol     = 3,
  pt.size  = 0.05,
  alpha    = 0.3
) & theme(axis.title.x = element_blank())
```

```{r}
#| label: qc-scatter
#| eval: false
#| fig-cap: "Scatter plots reveal relationships between QC metrics."

p1 <- FeatureScatter(seu, "nCount_RNA", "nFeature_RNA") +
  geom_hline(yintercept = c(200, 6000), linetype = "dashed", color = "#FF6B6B", linewidth = 0.6)

p2 <- FeatureScatter(seu, "nCount_RNA", "pct_mt") +
  geom_hline(yintercept = 20, linetype = "dashed", color = "#FF6B6B", linewidth = 0.6)

p1 | p2
```

---

## Apply Filters

```{r}
#| label: filter
#| eval: false

seu <- subset(
  seu,
  subset = nFeature_RNA > 200  &
           nFeature_RNA < 6000 &
           pct_mt < 20
)

cat("Cells retained:", ncol(seu), "\n")
```

::: {.callout-warning}
## Threshold Selection is Dataset-Specific
The values above are appropriate for PBMC datasets. Adjust based on:

- **Cell type composition** — neurons and hepatocytes naturally express more genes
- **Sequencing depth** — deeper libraries shift the `nCount_RNA` distribution
- **Tissue of origin** — immune cells have lower ambient RNA than solid tissues
:::

---

## Doublet Detection

Doublets are droplets containing two cells that appear as one. Even after ambient
RNA filtering, doublets can form distinct spurious clusters.

```{r}
#| label: doublet-normalization
#| eval: false

# DoubletFinder requires a preliminary normalization + PCA
# (these will be done properly in Chapter 02; here they are run solely for doublet detection)
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu)
seu <- ScaleData(seu)
seu <- RunPCA(seu, npcs = 20)
```

```{r}
#| label: doublet-finder
#| eval: false

library(DoubletFinder)

# Expected doublet rate: ~0.8% per 1,000 cells recovered (10x estimate)
nExp <- round(0.075 * ncol(seu))

seu <- doubletFinder(
  seu,
  PCs     = 1:20,
  pN      = 0.25,
  pK      = 0.09,   # pK should be optimised per sample — see ?paramSweep
  nExp    = nExp,
  reuse.pANN = FALSE
)

# Rename the classification column for clarity
seu$doublet_class <- seu[[grep("DF.classifications", colnames(seu@meta.data), value = TRUE)]]

# Remove doublets
seu <- subset(seu, subset = doublet_class == "Singlet")
cat("Singlets retained:", ncol(seu), "\n")
```

::: {.callout-tip}
## Optimising pK
Run `paramSweep()` followed by `summarizeSweep()` and `find.pK()` to determine
the optimal `pK` value for your dataset. This adds ~5–10 minutes but meaningfully
improves doublet detection accuracy.
:::

---

## Summary Statistics

```{r}
#| label: summary
#| eval: false

data.frame(
  Step       = c("After CreateSeuratObject", "After QC filter", "After doublet removal"),
  Cells      = c(ncol_raw, ncol_filtered, ncol(seu)),
  Median_nGene = c(
    median(raw_meta$nFeature_RNA),
    median(filt_meta$nFeature_RNA),
    median(seu$nFeature_RNA)
  )
)
```

---

## Save Checkpoint

```{r}
#| label: save
#| eval: false

saveRDS(seu, "data/processed/seu_01_filtered.rds")
```

---

## Session Info

```{r}
#| label: session-info
sessionInfo()
```
