---
title: "Normalization & Feature Selection"
subtitle: "Correct for sequencing depth and identify highly variable genes"
categories: [scRNA-seq, Seurat, Normalization]
---

<span class="chapter-badge">scRNA-seq · Chapter 02</span>

## Overview

Raw UMI counts are confounded by differences in sequencing depth across cells.
Normalization removes this technical variation, allowing meaningful comparison
of gene expression levels. Feature selection then focuses downstream analysis
on genes that carry the most biological signal.

**This chapter covers:**

1. Log-normalization (classical approach)
2. SCTransform v2 (variance-stabilizing approach)
3. Highly variable gene (HVG) selection
4. Data scaling

---

## Setup

```{r}
#| label: setup

library(Seurat)
library(ggplot2)
library(patchwork)
```

```{r}
#| label: load
#| eval: false

seu <- readRDS("data/processed/seu_01_filtered.rds")
```

---

## Approach 1: Log-Normalization

The standard approach: normalize each cell to 10,000 counts (CPM-like),
then apply a log1p transformation.

```{r}
#| label: lognorm
#| eval: false

seu <- NormalizeData(
  seu,
  normalization.method = "LogNormalize",
  scale.factor         = 10000
)
```

---

## Approach 2: SCTransform v2 (Recommended)

SCTransform uses regularized negative binomial regression to model
the mean-variance relationship, producing Pearson residuals that are
largely independent of sequencing depth.

```{r}
#| label: sctransform
#| eval: false

seu <- SCTransform(
  seu,
  vst.flavor          = "v2",         # use SCTransform v2
  vars.to.regress     = "pct_mt",     # optionally regress out mt percentage
  variable.features.n = 3000,
  verbose             = FALSE
)
```

::: {.callout-tip}
## Which normalization should I use?
- **LogNormalize** → faster, well-understood, compatible with all downstream Seurat functions
- **SCTransform v2** → generally better for datasets with high sequencing depth variance;
  required when integrating with `SCTransform`-normalized reference datasets

For most published workflows, SCTransform v2 is now the default recommendation.
:::

---

## Highly Variable Gene Selection

HVGs are genes that vary more across cells than expected by chance —
they capture biologically meaningful variation.

```{r}
#| label: hvg
#| eval: false

# If using LogNormalize (skip if using SCTransform — it selects HVGs internally)
seu <- FindVariableFeatures(
  seu,
  selection.method = "vst",
  nfeatures        = 3000
)

# Visualise top HVGs
top20 <- head(VariableFeatures(seu), 20)

p <- VariableFeaturePlot(seu)
LabelPoints(plot = p, points = top20, repel = TRUE, xnudge = 0, ynudge = 0)
```

---

## Scaling

Scaling centers each gene to mean 0 and variance 1 across cells,
preventing highly expressed genes from dominating PCA.

```{r}
#| label: scale
#| eval: false

# Scale only HVGs (faster; sufficient for PCA/UMAP)
seu <- ScaleData(seu, features = VariableFeatures(seu))

# --- OR scale all genes (required for heatmaps showing non-HVG genes) ---
seu <- ScaleData(seu, features = rownames(seu))
```

::: {.callout-note}
When using SCTransform, scaling is performed internally.
Calling `ScaleData()` again is not necessary for standard workflows.
:::

---

## Save Checkpoint

```{r}
#| label: save
#| eval: false

saveRDS(seu, "data/processed/seu_02_normalized.rds")
```

---

## Session Info

```{r}
#| label: session-info
sessionInfo()
```
