---
title: "Trajectory Analysis"
subtitle: "Infer developmental pseudotime with Monocle3"
categories: [scRNA-seq, Monocle3, Trajectory, Pseudotime]
---

<span class="chapter-badge">scRNA-seq · Chapter 07</span>

## Overview

Trajectory analysis orders cells along a continuum of developmental states,
assigning each cell a **pseudotime** value that reflects its progress along
a biological process (differentiation, activation, cell cycle, etc.).

**This chapter covers:**

1. Converting a Seurat object to a Monocle3 `cell_data_set`
2. Preprocessing and dimensionality reduction within Monocle3
3. Learning the principal graph
4. Setting the root node and computing pseudotime
5. Identifying genes that change along pseudotime

---

## Setup

```{r}
#| label: setup

library(Seurat)
library(monocle3)
library(ggplot2)
library(patchwork)
library(dplyr)
```

```{r}
#| label: load
#| eval: false

seu <- readRDS("data/processed/seu_05_annotated.rds")
```

---

## Convert Seurat → Monocle3 `cell_data_set`

```{r}
#| label: convert
#| eval: false

# Extract components from Seurat
gene_metadata <- data.frame(
  gene_short_name = rownames(seu),
  row.names       = rownames(seu)
)

cds <- new_cell_data_set(
  expression_data   = GetAssayData(seu, assay = "RNA", layer = "counts"),
  cell_metadata     = seu@meta.data,
  gene_metadata     = gene_metadata
)
```

---

## Preprocess and Reduce Dimensions

```{r}
#| label: preprocess
#| eval: false

cds <- preprocess_cds(cds, num_dim = 30)
cds <- reduce_dimension(cds, reduction_method = "UMAP")
```

::: {.callout-tip}
## Transfer UMAP coordinates from Seurat
To keep your Seurat UMAP layout consistent with the trajectory plot,
embed the Seurat UMAP coordinates directly into the Monocle3 object:

```r
reducedDim(cds, type = "UMAP") <- Embeddings(seu, reduction = "umap")
```
:::

---

## Cluster and Learn Graph

```{r}
#| label: cluster-graph
#| eval: false

cds <- cluster_cells(cds, resolution = 1e-3)

cds <- learn_graph(
  cds,
  use_partition     = TRUE,
  close_loop        = FALSE,
  learn_graph_control = list(minimal_branch_len = 5)
)

plot_cells(
  cds,
  color_cells_by = "cell_type",
  label_groups_by_cluster = FALSE,
  label_leaves   = TRUE,
  label_branch_points = TRUE
)
```

---

## Order Cells in Pseudotime

```{r}
#| label: pseudotime
#| eval: false

# Interactive root selection (run in RStudio, not on HPC)
cds <- order_cells(cds)

# --- OR set root programmatically ---
# Identify the root node nearest to a known progenitor cluster
root_cells <- rownames(subset(colData(cds), cell_type == "HSC"))
cds <- order_cells(cds, root_cells = root_cells)

plot_cells(
  cds,
  color_cells_by = "pseudotime",
  label_cell_groups = FALSE,
  label_leaves   = FALSE,
  label_branch_points = FALSE,
  graph_label_size = 3
)
```

---

## Identify Pseudotime-Dependent Genes

```{r}
#| label: deg-pseudotime
#| eval: false

# Test all genes (computationally intensive — subset to HVGs if needed)
pr_test <- graph_test(cds, neighbor_graph = "principal_graph", cores = 8)

pr_deg <- pr_test %>%
  filter(q_value < 0.05) %>%
  arrange(desc(morans_I))

head(pr_deg, 20)
```

```{r}
#| label: gene-modules
#| eval: false
#| fig-cap: "Co-expressed gene modules along the principal trajectory"

# Group co-expressed genes into modules
gene_module_df <- find_gene_modules(
  cds[rownames(pr_deg)[1:500], ],
  resolution = c(10^seq(-6, -1))
)

plot_cells(
  cds,
  genes        = gene_module_df,
  label_cell_groups = FALSE,
  show_trajectory_graph = FALSE
)
```

---

## Save

```{r}
#| label: save
#| eval: false

saveRDS(cds, "data/processed/cds_07_trajectory.rds")

# Add pseudotime back to Seurat object
seu$pseudotime <- pseudotime(cds)[colnames(seu)]
saveRDS(seu, "data/processed/seu_07_pseudotime.rds")
```

---

## Session Info

```{r}
#| label: session-info
sessionInfo()
```
