---
title: "Signac: scRNA + scATAC Integration"
subtitle: "Link chromatin accessibility to gene expression using Signac"
categories: [Multi-omics, Signac, scATAC-seq, scRNA-seq]
---

<span class="chapter-badge">Multi-omics · Chapter 02</span>

## Overview

[Signac](https://stuartlab.org/signac/) extends Seurat to handle chromatin
accessibility data. When paired with matched scRNA-seq (e.g., 10x Multiome),
it enables joint analysis — linking open chromatin regions to gene expression
at single-cell resolution.

**This chapter covers:**

1. Creating a Signac object from a 10x Multiome dataset
2. ATAC-seq QC within Signac
3. LSI dimensionality reduction
4. Joint WNN embedding with matched RNA
5. Linking peaks to genes

---

## Setup

```{r}
#| label: setup

library(Seurat)
library(Signac)
library(EnsDb.Hsapiens.v86)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ggplot2)
library(patchwork)
```

---

## Load 10x Multiome Data

```{r}
#| label: load-data
#| eval: false

# Read Cell Ranger ARC output
counts <- Read10X_h5("data/raw/pbmc_multiome/filtered_feature_bc_matrix.h5")

# Separate RNA and ATAC matrices
rna_counts  <- counts$`Gene Expression`
atac_counts <- counts$Peaks

# Load fragment file metadata
fragpath <- "data/raw/pbmc_multiome/atac_fragments.tsv.gz"
```

---

## Create Seurat + Signac Object

```{r}
#| label: create-object
#| eval: false

# Genome annotation
annotation <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)
seqlevels(annotation) <- paste0("chr", seqlevels(annotation))

# Create chromatin assay
atac_assay <- CreateChromatinAssay(
  counts     = atac_counts,
  sep        = c(":", "-"),
  fragments  = fragpath,
  annotation = annotation
)

# Build Seurat object with both modalities
seu <- CreateSeuratObject(counts = rna_counts, assay = "RNA")
seu[["ATAC"]] <- atac_assay

seu
```

---

## ATAC QC

```{r}
#| label: atac-qc
#| eval: false

DefaultAssay(seu) <- "ATAC"

seu <- NucleosomeSignal(seu)
seu <- TSSEnrichment(seu, fast = FALSE)

# Filter
seu <- subset(
  seu,
  subset = nCount_ATAC > 1000   &
           nCount_ATAC < 100000 &
           nCount_RNA  > 200    &
           nCount_RNA  < 25000  &
           nucleosome_signal < 2 &
           TSS.enrichment > 1
)
```

```{r}
#| label: atac-qc-plot
#| eval: false
#| fig-cap: "ATAC QC metrics: nucleosome signal and TSS enrichment"

VlnPlot(
  seu,
  features = c("nCount_ATAC", "TSS.enrichment", "nucleosome_signal"),
  ncol = 3, pt.size = 0
)
```

---

## Process Each Modality

```{r}
#| label: process-rna
#| eval: false

DefaultAssay(seu) <- "RNA"
seu <- SCTransform(seu, verbose = FALSE)
seu <- RunPCA(seu, npcs = 30)
```

```{r}
#| label: process-atac
#| eval: false

DefaultAssay(seu) <- "ATAC"
seu <- RunTFIDF(seu)
seu <- FindTopFeatures(seu, min.cutoff = "q0")
seu <- RunSVD(seu)   # LSI
```

---

## Joint WNN Embedding

```{r}
#| label: wnn
#| eval: false

seu <- FindMultiModalNeighbors(
  seu,
  reduction.list = list("pca", "lsi"),
  dims.list      = list(1:30, 2:50),  # exclude LSI component 1 (correlated with depth)
  modality.weight.name = "RNA.weight"
)

seu <- RunUMAP(
  seu,
  nn.name    = "weighted.nn",
  reduction.name = "wnn.umap",
  reduction.key  = "wnnUMAP_"
)

seu <- FindClusters(seu, graph.name = "wsnn", resolution = 0.8)
```

```{r}
#| label: wnn-plot
#| eval: false
#| fig-cap: "WNN UMAP colored by cluster and RNA modality weight"

p1 <- DimPlot(seu, reduction = "wnn.umap", label = TRUE) + NoLegend()
p2 <- FeaturePlot(seu, features = "RNA.weight", reduction = "wnn.umap")
p1 | p2
```

---

## Link Peaks to Genes

```{r}
#| label: link-peaks
#| eval: false

DefaultAssay(seu) <- "ATAC"

seu <- LinkPeaks(
  seu,
  peak.assay = "ATAC",
  expression.assay = "SCT",
  genes.use  = VariableFeatures(seu, assay = "SCT")
)
```

```{r}
#| label: coverage-plot
#| eval: false
#| fig-cap: "Coverage plot showing chromatin accessibility linked to MS4A1 (CD20)"

CoveragePlot(
  seu,
  region     = "MS4A1",
  features   = "MS4A1",
  expression.assay = "SCT",
  extend.upstream   = 500,
  extend.downstream = 10000
)
```

---

## Save

```{r}
#| label: save
#| eval: false

saveRDS(seu, "data/processed/seu_multiome_signac.rds")
```

---

## Session Info

```{r}
#| label: session-info
sessionInfo()
```
